---
title: "From a GBIF download to a defensible dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From a GBIF download to a defensible dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7.2, fig.height = 4.2, dpi = 110
)
```

```{r setup}
library(ingrain)
```

## The question the audit answers

A GBIF download carries a per-record positional uncertainty radius,
`coordinateUncertaintyInMeters`, reported for some records and missing
for others. Coordinate-cleaning tools flag impossible and suspicious
coordinates; what they do not say is what the *reported uncertainty*
means for the analysis you are about to run. A widespread workflow idiom
keeps records whose radius is below a threshold *or* missing, and drops
the rest -- a filter that selects on which publishers reported metadata,
not on where the organisms were.

`ingrain` asks a different question, record by record: **at your
analysis resolution, can an uncertainty-aware treatment act on this
record at all?** For a raster with cell edge length *R* and a reported
radius *r*, every record lands in one of four states:

* **inert** (*r* &le; *R*/2) -- the uncertainty disc fits within one
  cell, so corrections cannot meaningfully act; the covariate is
  effectively constant over the region the record could occupy.
* **marginal** (*R*/2 < *r* &le; 3*R*) -- the boundary band, where the
  effect of positional error depends on where the record sits relative
  to cell edges and on local covariate variation.
* **actionable** (*r* > 3*R*) -- the uncertainty region spans many
  cells; treatments have room to act, and the choice among them has
  consequences.
* **unreported** (*r* is `NA`) -- no treatment can act, and the radius
  cannot be recovered from the coordinates themselves.

## One grain

The bundled `crayfish` sample is a fixed 5,000-record draw from a
real GBIF download (see the data note at the end). At a conventional
1 km grain:

```{r bar, fig.height = 2.3}
a <- ingrain(crayfish, grain = 1000)
a
autoplot(a)
```

More than half of these records are inert: whatever the merits of the
available treatments, they cannot change the model *for those records
at this resolution*. Hefley et al. (2017) observed in passing that a
change-of-support correction cannot act where the covariate is constant
within a grid cell; the `inert` state turns that observation into a
per-record screen, run before any model is fitted.

## The grain is a choice

The partition is not a property of the data alone -- it is a property
of the data *and* the resolution. Sweeping the grain makes that
explicit:

```{r profile}
p <- ingrain_profile(crayfish, mark = 1000)
p
autoplot(p)
```

As the grain coarsens, records migrate from actionable into inert; the
treatment debate concerns a shrinking share of the data. The grey band
does not move: no choice of resolution touches the unreported share.
The steps in the coloured bands are real, not artefacts -- reported
radii cluster at publisher conventions (301 m, 1 km, 5 km, ...), and
the classification thresholds cross those mass points as the grain
sweeps.

## Publishers, not organisms

Cutting the same audit by publishing dataset shows where the partition
comes from:

```{r datasets, fig.height = 6}
bd <- by_dataset(a)
bd
autoplot(bd)
```

Most datasets sit almost entirely in a single state -- publishers
occupy regimes, not a gradient. The uncertainty coefficient quantifies
this: how much does knowing the dataset reduce your uncertainty about a
record's state?

```{r u}
ingrain_u(a, seed = 1)
```

Read against its permutation null, the excess is what survives having
many groups. The practical consequence: you cannot predict the
partition of your own download from the taxon; datasets within one
taxon differ as much as taxa do. Audit your download, not your
organism.

## What to do with each state

* **inert** -- use the records as they are, at this grain. Corrections
  cannot act on them; discarding them for their radius would be pure
  loss.
* **marginal** -- treat as a sensitivity band: small changes of grain
  move these records across the boundary, so check that conclusions do
  not hinge on them.
* **actionable** -- this is where method choice matters. Options
  include threshold filtering, the near-geographic and
  near-environmental point methods of Smith et al. (2023) implemented
  in `enmSdmX`, and change-of-support estimation (Hefley et al. 2017).
* **unreported** -- do not impute a radius from coordinate decimal
  places: decimals bound the uncertainty only from below, so they
  cannot supply the missing upper bound. For records with gridded
  coordinates -- a substantial share of GBIF (Feng et al. 2023) --
  GridDER can recover a grid-implied radius; for the rest, the missing
  radius is irreducible, and the honest options are to analyse with and
  without these records or to model reporting explicitly.

## What ingrain is not

`ingrain` deliberately performs no cleaning, no imputation and no
estimation of radii. Run coordinate cleaning first (e.g.
CoordinateCleaner, Zizka et al. 2019; bdc, Ribeiro et al. 2022), then
the audit, then whatever treatment the actionable share justifies.

## A note on the bundled data

`crayfish` is drawn from GBIF occurrence download
[doi:10.15468/dl.99ezk2](https://doi.org/10.15468/dl.99ezk2) and
restricted to CC0/CC BY records so it can be redistributed. Because
record licences are set by the publishing dataset, that restriction is
itself a publisher filter, and the shares above differ by construction
from the unfiltered download. The sample demonstrates the mechanics;
audit your own download for your own numbers.

## References

Feng, X., et al. (2023) GridDER: Grid Detection and Evaluation in R.
*Ecological Informatics*.

Hefley, T.J., Brost, B.M. & Hooten, M.B. (2017) Bias correction of
bounded location errors in presence-only data. *Methods in Ecology and
Evolution*, 8, 1566--1573.

Marcer, A., et al. (2022) Quality issues in georeferencing: from
physical collections to digital data repositories for ecological
research. *Ecography*, 2022, e06025.

Moudrý, V., et al. (2024) Optimising occurrence data in species
distribution models. *Ecography*, 2024, e07294.

Ribeiro, B.R., et al. (2022) bdc: A toolkit for standardizing,
integrating and cleaning biodiversity data. *Methods in Ecology and
Evolution*, 13, 1421--1428.

Smith, A.B., Murphy, S.J., Henderson, D. & Erickson, K.D. (2023)
Including imprecisely georeferenced specimens improves accuracy of
species distribution models and estimates of niche breadth. *Global
Ecology and Biogeography*, 32, 342--355.

Zizka, A., et al. (2019) CoordinateCleaner: Standardized cleaning of
occurrence records from biological collection databases. *Methods in
Ecology and Evolution*, 10, 744--751.
